---
title: "Urban FIA community ecology analysis"
author: "Clifton McKee"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

## Load packages and functions

```{r packages-functions}
# Load packages
library(tidyverse)
library(readxl)
library(janitor)
library(sf)
library(vegan)
library(ggrepel)
```

### Data import

* Data sources:
  + [EPA Ecoregions of North America](https://www.epa.gov/eco-research/ecoregions-north-america)
  + [United States Cities Database](https://simplemaps.com/data/us-cities)

```{r data-import}
# Import main urban tree data file and clean some columns
trees <- read_csv("Data/cities_2022_trees.csv") %>% 
  rename(RowID = `...1`) %>% 
  mutate(city = str_sub(EVALID, 1, -9),
         city = case_when(city == "MinneapolMN" ~ "MinneapolisMN",
                          city == "SpringfielMO" ~ "SpringfieldMO",
                          TRUE ~ city))

# Import city land use data file and clean some columns
landuse <- read_excel("Data/cities_2022_ownership_landuse.xlsx",
                      sheet = "Cities") %>% 
  mutate(city = case_when(city == "MinneapolMN" ~ "MinneapolisMN",
                          city == "SpringfielMO" ~ "SpringfieldMO",
                          TRUE ~ city))

# Import EPA Ecoregions level I data
ecoregions1 <- st_read("Data/Ecoregions/na_cec_eco_l1/NA_CEC_Eco_Level1.shp",
                       quiet = TRUE) %>% 
  mutate(NA_L1NAME = str_to_lower(NA_L1NAME))

# Import database of US cities with coordinates
city_coords <- read_csv("Data/City coordinates/uscities.csv") %>% 
  mutate(camelname = paste0(str_remove_all(city, "[\\.\\s]"), state_id))

# Filter the city coordinates to the cities with urban tree data
tree_city_coords <- city_coords %>% 
  filter(camelname %in% landuse$city)

# Merged city data with coordinates
tree_city_merge <- full_join(x = landuse,
                             y = tree_city_coords,
                             by = c("city" = "camelname"))
```

### City profiles

#### Number of plots

```{r city-plots}
# Plot total city plots
tree_city_merge %>% 
  ggplot() +
  geom_col(aes(x = reorder(city, lng, decreasing = FALSE),
               y = plot_count_Curr)) +
  labs(x = "City (ordered by longitude)", y = "Total city plots") +
  theme_bw(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### Number of forested plots

```{r forested-plots}
# Forested plots per city
trees %>% 
  full_join(tree_city_coords, by = c("city" = "camelname")) %>%
  group_by(city, lat, lng) %>%
  summarize(plots = n_distinct(PLOTID)) %>% 
  ggplot() +
  geom_col(aes(x = reorder(city, lng, decreasing = FALSE),
               y = plots)) +
  labs(x = "City (ordered by longitude)", y = "Total forested city plots") +
  theme_bw(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### Total land area

```{r city-acreage}
# Plot total city land area
tree_city_merge %>% 
  ggplot() +
  geom_col(aes(x = reorder(city, lng, decreasing = FALSE),
               y = city_acres)) +
  labs(x = "City (ordered by longitude)", y = "Total city land area (acres)") +
  theme_bw(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### Total population

```{r city-population}
# Plot total city population
tree_city_merge %>% 
  ggplot() +
  geom_col(aes(x = reorder(city, lng, decreasing = FALSE),
               y = population)) +
  labs(x = "City (ordered by longitude)", y = "Population") +
  theme_bw(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### Population density

```{r city-density}
# Plot city population density
tree_city_merge %>% 
  ggplot() +
  geom_col(aes(x = reorder(city, lng, decreasing = FALSE),
               y = density)) +
  labs(x = "City (ordered by longitude)", y = "Population density (per sq km)") +
  theme_bw(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### Ownership of city land area

```{r land-ownership}
# Plot the proportion of city acreage by ownership
tree_city_merge %>% 
  pivot_longer(ForestService:StateLocal,
               names_to = "Ownership",
               values_to = "Acreage") %>% 
  ggplot() +
  geom_col(aes(x = reorder(city, lng, decreasing = FALSE),
               y = Acreage,
               fill = Ownership),
           position = "fill") +
  scale_fill_viridis_d(option = "H") +
  labs(x = "City (ordered by longitude)", y = "Proportion of city land area") +
  theme_bw(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### City land area by land use type

```{r land-use}
# Plot the proportion of city acreage by landuse type
tree_city_merge %>% 
  pivot_longer(Agriculture:Water,
               names_to = "Type",
               values_to = "Acreage") %>% 
  ggplot() +
  geom_col(aes(x = reorder(city, lng, decreasing = FALSE),
               y = Acreage,
               fill = Type),
           position = "fill") +
  scale_fill_viridis_d(option = "H") +
  labs(x = "City (ordered by longitude)", y = "Proportion of city land area") +
  theme_bw(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### Ecoregions of cities

```{r ecoregions1}
# Simplify the polygons
ecoregions1_simple <- st_simplify(ecoregions1, preserveTopology = TRUE, dTolerance = 1000) %>% 
  st_transform(crs = "EPSG:4326")

# Turn coordinates into an sf object with correct CRS
tree_city_sf <- st_as_sf(tree_city_coords, coords = c("lng", "lat"), crs = st_crs(ecoregions1_simple))

# Plot the cities over ecoregions (level I)
ggplot() +
  geom_sf(data = ecoregions1_simple, aes(fill = NA_L1NAME)) +
  geom_sf(data = tree_city_sf) +
  scale_fill_viridis_d(name = "Ecoregion I", option = "H") +
  theme_void(base_size = 8) +
  theme(legend.position = "bottom") +
  coord_sf(xlim = c(-125, -67), ylim = c(24, 49))
```

### Species diversity in surveyed cities

#### Species accumulation curves

```{r species-accumulation}
# Convert tree data to community dataset, pooled by plots within city
tree_community_plots <- trees %>% 
  group_by(city, PLOTID, scientific) %>% 
  summarize(count = n()) %>% 
  pivot_wider(names_from = scientific, values_from = count) %>% 
  replace(is.na(.), 0)

# Calculate species accumulation curves across cities
poolaccum_fun <- function(city, nperm, measure) {
  sp <- poolaccum(tree_community_plots[tree_community_plots$city == city, -c(1, 2)],
                  permutations = nperm, minsize = 3)
  data.frame(city = city, summary(sp)[[measure]])
}
poolaccum_S <- NULL
for(i in 1:length(tree_city_merge$city)){
  poolaccum_S <- rbind(poolaccum_S, poolaccum_fun(tree_city_merge$city[i], 100, "S"))
}
poolaccum_S <- clean_names(poolaccum_S)

# Plot species accumulation curves by city
poolaccum_S %>% 
  ggplot() +
  geom_ribbon(aes(x = n, ymin = x2_5, ymax = x97_5, fill = city), alpha = 0.1) +
  geom_line(aes(x = n, y = s, color = city)) +
  geom_text_repel(data = poolaccum_S %>% 
              group_by(city) %>% 
              slice_tail(n = 1),
              aes(label = paste0(city, " = ", s),
                  x = n,
                  y = s,
                  color = city),
              nudge_x = 5,
              min.segment.length = 2,
              size = 2.5) +
  scale_color_viridis_d(option = "H") +
  scale_fill_viridis_d(option = "H") +
  labs(x = "Number of forested plots", y = "Tree species") +
  theme_bw(base_size = 10) +
  theme(legend.position = "none")
```

#### Extrapolated species richness

```{r extrapolate-species}
# Extrapolate species diversity from pooled plots within cities
pooled_accum <- with(tree_community_plots[, c(1, 2)],
             specpool(tree_community_plots[, -c(1, 2)],
                      city,
                      smallsample = TRUE)) %>%
  rownames_to_column("city") %>% 
  rename(Species.est = Species,
         chao.est = chao,
         jack1.est = jack1,
         jack2.est = jack2,
         boot.est = boot) %>% 
  pivot_longer(Species.est:boot.se,
               names_to = c("estimate", "measure"), names_sep = "\\.",
               values_to = "value") %>% 
  pivot_wider(names_from = measure, values_from = value)

# Plot extrapolated species richness indices by city
pooled_accum %>% 
  ggplot() +
  geom_col(aes(x = city, y = est)) +
  geom_errorbar(aes(x = reorder(city, est, median, decreasing = TRUE), ymin = est - se, ymax = est + se), width = 0) +
  facet_wrap(~estimate) +
  labs(x = "City (ordered by median richness)", y = "Species richness estimate") +
  theme_bw(base_size = 10) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 8))
```
